Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints

### 2014-06-03 Compare_053014_statsmed_revisions.tex
edit

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating univariate biomarker as a PS of continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

We propose the use of several curves for displaying surrogate quality which consider treatment efficacy within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves. Evaluation of multivariate principal surrogates is complicated by the increased dimension.  Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect. The strength of this association can be displayed by the CEP curve \citep{Gilbert08}. The CEP is a contrast of risk over the treatment arms within subgroups defined by the candidate surrogate under active treatment and marginalizing over the candidate surrogate under control. The dimension of the marginal CEP function increases with the dimension of the candidate surrogate, making the marginal CEP difficult to display and interpret for multivariate candidate PS. This makes the curves two-dimensional regardless of the number of biomarkers included in the risk model. Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define. We also consider the cumulative distribution function (CDF) of the marginal CEP function. Time-dependent TPF and FPF curves based on the levels of diagnostic tests and time-independent curves based on levels of causal contrasts, \citep{Huang12b}, have been proposed in previous works. However, no version of the TPF or FPF curves have been proposed for the evaluation or comparison of candidate PSs, nor has the causal-contrast-based and time-dependent versions of the curves been defined.

We develop summary statistics for the comparison of candidate PS for a continuous time-to-event end point. Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS . Often evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality. We develop two extensions of the standard total gain \citet{Huang11}, both of which provide a single values measure of surrogate quality that can be used to compare candidate surrogates for the same clinical output in the same trial. The time-dependent standard total gain, ($STG(t)$), allows for a single value comparison of marginal CEP models as well as inference for comparison of candidate PS that summaries all information and properly accounts for censoring at and given time point $t$. Over a range of time points, a display of $STG(t)$ versus $t$ curve provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time. The time integrated  standard total gain, ($widetilde{STG}$), which is the time-dependent standard total gain integrated over the estimated survive times, provides a single value summary of surrogate quality independent of time.

In Section 2, we outline our notation. In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PSs or the comparison of candidate PSs.  In Section 3.1, we describe the time-dependent risk estimands and outline the assumptions that are helpful for identifying these estimands.  We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.  We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison. In Section 3.7 we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}. In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PSs. In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT). We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combination of the two candidates is not an improvement over the higher quality surrogate alone. In Section 6, we discuss potential limitations of our methods and future avenues for research.

@@

All $S_j$ are measured at the same fixed time point $\tau$ after randomization, and this time point, post-randomization but prior to clinical-event, and z define the surrogate measurement. If $T_{j,i}(z)$ is less than $\tau$ for a given subject $i$, all $S_{j,i}(z)$ are undefined for that subject, as observation of the post-randomization pre-clinical-event measures $S_{j}(z)$ are no longer possible under either treatment condition. For this reason, subjects with min$(T,C) < \tau$ are excluded from the evaluation cohort. Let $M_j$ be the indicator that $S_{j}(1)$ is observed, and let $s_{1,j,i}$ be the realization of $S_{j}(1)$ for subject $i$. We assume the full set of observed and counterfactual outcomes are independently and identically distributed over the trial subjects $i= 1, \ldots n$. Let $F_{S_1(1), \ldots, S_J(1)}(\cdot,\dots,\cdot)$ and $F_{S_1(1), \ldots, S_J(1)|W}(\cdot|\cdot)$ be the joint CDF of the $S_1(1), \ldots, S_J(1)$ and joint CDF of the $S_1(1), \ldots, S_J(1)$ conditional on $W$, respectively. Define $\rho_z(t)=Pr(Y^t(z)=1)=E(Y^t(z))=Pr(T(z)\leq t)$ to be the marginal prevalence of the potential clinical outcome at or before time $t$ for treatment arm $z$. Our risk estimands of interest, $risk_z(s)$, are functions of the candidate surrogate that give some measure of risk under the treatment $z$.

\citet{Frangakis02} developed the concept of principal surrogacy and defined a PS by a criterion based on joint risk estimands, which conditions on the candidate surrogate under active treatment and under control, $risk_z(s_1,s_0)$. If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN). Subsequent work added a second criterion that a good PS should modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, what \citet{Gilbert08} called average causal sufficiency ACS. This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.  ACN is a strict but sensible necessary condition, which guarantees that there is no effect on the clinical out when there is no effect on the PS. However, it is clearly not a sufficient condition, as any biomarker in a trial with no clinical treatment effect is likely to satisfy ACN.  The combination of ACS and ACN are often looked evaluated in the current literature, with some works relaxing strict ACN.  ACS also suggests a possibly ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate. Strict ACN alone classify candidates as either PS or not PS, with no other indication of quality.

@@

For a univariate candidate PS strict or relaxed ACN can be displayed using the joint CEP or the marginal CEP.  Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically. As well, the definition does not seem as sensible in the multi-dimensional case. Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates. However, it seems like you would want less clinical treatment effect if there was no treatment effect on any of the biomarkers.  We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional of a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$. For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections 3.3 and 3.4. Like ACN, the classification accuracy measures allow for the quantification of the agreement between treatment effect on the clinical outcome and predictive model when there is no treatment effect on any of the biomarkers in the model.  Unlike ANC, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.  These accuracy measures can also be displayed as curves vs. the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model. We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary for a set of biomarkers to have value as a PS, but that this is not sufficient. Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.

ACS for a multivariate PS is wide variation in the CEP over subgroups defined by the set of biomarkers. This is difficult to display graphically due to the increased dimension of the surrogate. We propose the use of functions based on the CEP rather than the set of biomarker values for display of ACS. The propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS. We also suggest a summary statistic standardized total gain that quantifies both our suggested accuracy concept and ACS in a single measurement. Details for standard total gain are given in Section \ref{STG}.  As well, if one assumes a parametric a model for risk conditional on a set of biomarkers a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.

@@

\section{Methods}
\subsection{Time-Dependent Causal Risk Estimands} \label{CRE}
In the time-to-event setting there are many ways to define risk. The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF. The conditional CDF-based risk estimands can by defined as:
\begin{eqnarray*}
risk_{1}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_1(t|s_{1,1}, \ldots, s_{1,J})\\
&\equiv& P\{T(1) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{1}(1)=s_{1,J},T(1)> \tau, T(0)> \tau\},\\
risk_{0}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_0(t|s_{1,1}, \ldots, s_{1,J}) \\
&\equiv& P\{T(0) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{J}(1)=s_{1,J},T(1)>\tau, T(0)>\tau\}.
\end{eqnarray*}
The subscript of a function indicates the value of $z$. The comparison between the risk estimands over the treatment arms is the causal comparison of interest for principal surrogate evaluation. Our contrast of interest is defined as $\Delta(t|s_{1,1}, \dots, s_{1,J}) \equiv risk_0(t|s_{1,1}, \ldots, s_{1,J})-risk_1(t|s_{1,1}, \ldots, s_{1,J})$, the risk difference. Using the CDF based risk estimands, this contrast is causal because both risks condition on the same sets. This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}. For this reason, we focus on the CDF based marginal risks, which provide well defined  interpretations for our suggested summary statistics.

In order to link these risk estimands to the observable data we must make some assumptions; Assumptions A1-A4 reduce the number of missing potential outcomes and help identify the risk estimands.
\begin{itemize}
\item A1: Stable Unit Treatment Value Assumption (SUTVA) and Consistency 
\item A2: Ignorable Treatment Assignment
\item A3: Equal individual clinical risk up to time $\tau$: $T(1)\leq\tau$ if and only if $T(0)\leq\tau$
\item A4: Random censoring: $T(z) \perp C(z)$ for $z=\{0,1\}$.
\end{itemize} 
Assumption A1 implies that subject's potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency. For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$. SUTVA may not hold is some situations, such as cluster randomized vaccine trials where herd immunity may change the treatment effect on all outcomes in the cluster.  Assumption A2 implies that the distribution of the potential outcomes S(z) and Y(z) are independent of the observed treatment assignment. Assumption A2, is generally true in randomized, blinded trials.  Assumption A3 aides in the identification of subjects with min$(T,C) < \tau$ to be excluded from the evaluation cohort without causing bias. Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced events or drop-out over the trial arms prior to $\tau$. Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used. Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z=\{0,1\}$.

@@

One can test the null hypothesis of no surrogate value or no ACS, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$. If any of the $S_{J}(1)$ associated terms significantly differ from zero there is evidence of variation in the risk difference over the candidate PS.

@@

\subsection{Cumulative Distribution Function of $\Delta$} \label{CDF}
 Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PSs and for the comparison of candidate PS. An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$. If we assume monotonicity of treatment effect, formally stated as
\begin{itemize}
\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z=\{0,1\}$,
\end{itemize}
then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$). If we alternatively assume no benefit, then the line will be at the proportion of subjects with $D(t)=-1$. Assumption A6 is not needed for the function $CDF^{t}_{\Delta}(c)$ to be well-defined. For a useless surrogate, the plot of $CDF^{t}_{\Delta}(c)$ versus $c$ will be a vertical line at $\rho_0(t)-\rho_1(t)$, the difference in prevalence of the potential clinical outcomes between the trial arms. Partially useful surrogates will have $CDF^{t}_{\Delta}(c)$ versus $c$ curves between these two curves with better surrogates having flatter curves. Figure \ref{4panel} displays the CDF of the risk difference for several different surrogate quality levels. We suggest the plug-in estimator of $\widehat{CDF}^{t}_{\Delta}(c)$, $\widehat{CDF}^{t}_{\Delta}(c)=\int{\mathrm{I}_{[\hat{\Delta}(t|s_{1,1,i}, \dots, s_{1,j,i})\leq c]} d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}$. Although we display a useless surrogate in Figure \ref{4panel}, we suggest testing for any surrogate value prior to estimation of $\widehat{CDF}^{t}_{\Delta}(c)$ as it was found in simulations that there is evidence of bias when there is no variation in risk difference; a suggested test is given in Section \ref{CRE}.

@@

We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$. We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $STG(3)$ of 0.23 and 0.51, respectively. We investigate these different levels of candidate surrogates for two different BIP correlations, $R_{s_{1j}w}$, of 0.8 and 0.5. For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0.8), (0.5, 0.5)\}$ with a correlation between the candidate surrogates ranging from 0 to 0.3. Results for the 0.3 correlation are displayed. 500 simulation were run for each scenario; some simulations did not converge and were removed from the simulations used for bias and power calculations, there was a 2.5\% non-convergence rate  All candidate PS in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.

>>>>

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating univariate biomarker as a PS of continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves. A CEP is any function of the treatment arm specific biomarker conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero. A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement of $S$. Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect and the strength of this association can be displayed by the marginal-CEP versus candidate PS under active treatment figure \citep{Gilbert08}. Evaluation of multivariate principal surrogates is complicated by the increased dimension.

The dimension of the marginal-CEP versus candidate PS under active treatment figure increases with the dimension of the candidate surrogate, making the marginal-CEP difficult to display and interpret for multivariate candidate PS. If instead we consider the clinical outcome within level of the marginal-CEP we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model. Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define. We also consider the cumulative distribution function (CDF) of the marginal CEP function. Time-dependent TPF and FPF curves based on the levels of diagnostic tests and time-independent curves based on levels of causal contrasts, \citep{Huang12b}, have been proposed in previous works. However, no version of the TPF or FPF curves have been proposed for the evaluation or comparison of candidate PS, nor has the CEP-based and time-dependent versions of the curves been defined.

We develop summary statistics for the comparison of candidate PS for a continuous time-to-event endpoint. Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS . Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality. We develop two extensions of the standardized total gain \citet{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial. The time-dependent standardized total gain, ($STG(t)$), allows for the comparison of candidate PS at a given time point $t$ that properly accounts for censoring. Over a range of time points, a display of $STG(t)$ versus $t$ provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time. The time integrated  standardized total gain, ($\widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survival times, provides a single value summary of surrogate quality independent of time.

In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PS or the comparison of candidate PS.  In Section 3.1, we describe the time-dependent risk estimands and outline the assumptions that are helpful for identifying these estimands.  We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.  We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison. In Section 3.7 we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}. In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PS. In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT). We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combination of the two candidates is not an improvement over the higher quality surrogate alone. In Section 6, we discuss potential limitations of our methods and future avenues for research.

@@

All $S_j$ are measured at the same fixed post-randomization time point $\tau$. If $T_{j,i}(z)$ is less than $\tau$ for a given subject $i$, all $S_{j,i}(z)$ are undefined for that subject, as observation of the post-randomization pre-clinical-event measures $S_{j}(z)$ are no longer possible under either treatment condition. For this reason, subjects with $T< \tau$ are excluded from the evaluation cohort. Let $M_j$ be the indicator that $S_{j}(1)$ is observed, and let $s_{1,j,i}$ be the realization of $S_{j}(1)$ for subject $i$. We assume the full set of observed and counterfactual outcomes are independently and identically distributed over the trial subjects $i= 1, \ldots n$. Let $F_{S_1(1), \ldots, S_J(1)}(\cdot,\dots,\cdot)$ and $F_{S_1(1), \ldots, S_J(1)|W}(\cdot|\cdot)$ be the joint CDF of the $S_1(1), \ldots, S_J(1)$ and joint CDF of the $S_1(1), \ldots, S_J(1)$ conditional on $W$, respectively. Define $\rho_z(t)=Pr(Y^t(z)=1)=E(Y^t(z))=Pr(T(z)\leq t)$ to be the marginal prevalence of the potential clinical outcome at or before time $t$ for treatment arm $z$. Our risk estimands of interest, $risk_z(s)$, are functions of the candidate surrogate that give some measure of risk under the treatment $z$.

\citet{Frangakis02} developed the concept of principal surrogacy and defined a PS by a criterion based on joint risk estimands, which conditions on the candidate surrogate under active treatment and under control, $risk_z(s_1,s_0)$. If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN). Subsequent work added a second criterion that a good PS should modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, what \citet{Gilbert08} called average causal sufficiency ACS. This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.  ACN is a strict but sensible necessary condition, which guarantees that there is no effect on the clinical outcomewhen there is no effect on the PS. However, it is clearly not a sufficient condition, as any biomarker in a trial with no clinical treatment effect is likely to satisfy ACN.  The combination of ACS and ACN are often used for evaluation in the current literature, with some works relaxing strict ACN.  ACS also suggests a possibly ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate. Strict ACN alone classify candidates as either PS or not PS, with no other indication of quality.

@@

For a univariate candidate PS strict or relaxed ACN can be displayed using the joint CEP or the marginal CEP.  Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically. The definition does not seem as sensible in the multi-dimensional case. Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates. However, it seems like you would want less clinical treatment effect if there was no treatment effect on any of the biomarkers.  We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional of a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$. For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections 3.3 and 3.4. Like ACN, the classification accuracy measures allow for the quantification of the agreement between treatment effect on the clinical outcome and predictive model when there is no treatment effect on any of the biomarkers in the model.  Unlike ANC, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.  These accuracy measures can also be displayed as curves vs. the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model. We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary for a set of biomarkers to have value as a PS, but that this is not sufficient. Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.

ACS for a multivariate PS is wide variation in the CEP over subgroups defined by the set of biomarkers. This is difficult to display graphically due to the increased dimension of the surrogate. We propose the use of functions based on the CEP rather than the set of biomarker values for display of ACS. The propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS. We also suggest a summary statistic standardized total gain that quantifies both our suggested accuracy concept and ACS in a single measurement. Details for standardized total gain are given in Section \ref{STG}.  As well, if one assumes a parametric a model for risk conditional on a set of biomarkers, a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.

@@

\section{Methods}
\subsection{Time-Dependent Causal Risk Estimands} \label{CRE}
In the time-to-event setting there are many ways to define risk. The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF. The conditional CDF-based risk estimands can be defined as:
\begin{eqnarray*}
risk_{1}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_1(t|s_{1,1}, \ldots, s_{1,J})\\
&\equiv& P\{T(1) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{1}(1)=s_{1,J},T(1)> \tau, T(0)> \tau\},\\
risk_{0}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_0(t|s_{1,1}, \ldots, s_{1,J}) \\
&\equiv& P\{T(0) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{J}(1)=s_{1,J},T(1)>\tau, T(0)>\tau\}.
\end{eqnarray*}
The subscript of a function indicates the value of $z$. The comparison between the risk estimands over the treatment arms is the causal comparison of interest for principal surrogate evaluation. Our contrast of interest is defined as $\Delta(t|s_{1,1}, \dots, s_{1,J}) \equiv risk_0(t|s_{1,1}, \ldots, s_{1,J})-risk_1(t|s_{1,1}, \ldots, s_{1,J})$, the risk difference. Using the CDF based risk estimands, this contrast is causal because both risks condition on the same sets. This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}. For this reason, we focus on the CDF based marginal risks, which provide well defined  interpretations for our suggested summary statistics.

In order to link these risk estimands to the observable data we must make some assumptions; Assumptions A1-A4 reduce the number of missing potential outcomes and help identify the risk estimands.
\begin{itemize}
\item A1: Stable Unit Treatment Value Assumption (SUTVA) and Consistency 
\item A2: Ignorable Treatment Assignment
\item A3: Equal individual clinical risk up to time $\tau$: $T(1)\leq\tau$ if and only if $T(0)\leq\tau$
\item A4: Random censoring: $T(z) \perp C(z)$ for $z=\{0,1\}$.
\end{itemize} 
Assumption A1 implies that subjects' potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency. For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$. SUTVA may not hold is some situations, such as cluster randomized vaccine trials where herd immunity may change the treatment effect on all outcomes in the cluster.  Assumption A2 implies that the distribution of the potential outcomes S(z) and Y(z) are independent of the observed treatment assignment. Assumption A2, is generally true in randomized, blinded trials.  Assumption A3 aides in the identification of subjects with min$(T,C) < \tau$ to be excluded from the evaluation cohort without causing bias. Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced events or drop-out over the trial arms prior to $\tau$. Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used. Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z=\{0,1\}$.

@@

One can test the null hypothesis of no surrogate value or no ACS, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$. If any of the $S_{J}(1)$ associated terms significantly differ from zero then there is evidence of variation in the risk difference over the candidate PS.

@@

\subsection{Cumulative Distribution Function of $\Delta$} \label{CDF}
 Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PS and for the comparison of candidate PS. An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$. If we assume monotonicity of treatment effect, formally stated as
\begin{itemize}
\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z=\{0,1\}$,
\end{itemize}
then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$). If we alternatively assume no benefit, then the line will be at the proportion of subjects with $D(t)=-1$. Assumption A6 is not needed for the function $CDF^{t}_{\Delta}(c)$ to be well-defined. For a useless surrogate, the plot of $CDF^{t}_{\Delta}(c)$ versus $c$ will be a vertical line at $\rho_0(t)-\rho_1(t)$, the difference in prevalence of the potential clinical outcomes between the trial arms. Partially useful surrogates will have $CDF^{t}_{\Delta}(c)$ versus $c$ curves between these two curves with better surrogates having flatter curves. Figure \ref{4panel} displays the CDF of the risk difference for several different surrogate quality levels. We suggest the plug-in estimator of $\widehat{CDF}^{t}_{\Delta}(c)$, $\widehat{CDF}^{t}_{\Delta}(c)=\int{\mathrm{I}_{[\hat{\Delta}(t|s_{1,1,i}, \dots, s_{1,j,i})\leq c]} d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}$. Although we display a useless surrogate in Figure \ref{4panel}, we suggest testing for any surrogate value prior to estimation of $\widehat{CDF}^{t}_{\Delta}(c)$ as it was found in simulations that there is evidence of bias when there is no variation in risk difference; a suggested test is given in Section \ref{CRE}.

@@

We investigate four different univariate candidate PS with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$. We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $STG(3)$ of 0.23 and 0.51, respectively. We investigate these different levels of candidate surrogates for two different BIP correlations, $R_{s_{1j}w}$, of 0.8 and 0.5. For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0.8), (0.5, 0.5)\}$ with a correlation between the candidate surrogates ranging from 0 to 0.3. Results for the 0.3 correlation are displayed. 500 simulation were run for each scenario; some simulations did not converge and were removed from the simulations used for bias and power calculations, there was a 2.5\% non-convergence rate  All candidate PS in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.
